---
title: "Taxicab Project"
author: "Jack Dzialo, Tayten Bennetsen, Rith Katikaneni, Griffin Ball"
date: '10/1/24'
format: pdf
pdf: default
html: default
pdf-engine: pdflatex
editor: visual
fontsize: 10pt
geometry: margin=1in
toc: false 
toc-depth: 2 
---

# Introduction

Our team decided to analyze a dataset with information about taxi rides in Chicago to try and find links between taxi data such as ride time, price, and length with outside information such as weather conditions, different locations and their factors, and amount of traffic. The dataset contains over 5 million rows, and contains information of almost all taxi rides in the Chicago area from January 2024 to the present (October). It contains information such as but limited to: trip start and end times, trip length, fares and tips, company, and payment type.

# Plots

First, let us load in the data.

```{r}
library(data.table)
data <- data.table(fread("~/Downloads/taxi_chicago.csv"))
```

## Plot 1 - Ride Price vs Ride Time

#### Scatterplot

This plot shows price of ride price vs. ride time.

```{r}
indexes <- sample(1:10000, 1000)
ride_time <- data$`Trip Seconds`[indexes]
ride_price <- data$Fare[indexes]
outliers1 <- which(ride_price <= quantile(ride_price, .97, na.rm = TRUE))
outliers2 <- which(ride_time <= quantile(ride_time, .97, na.rm=TRUE))
threshold <- intersect(outliers1, outliers2)
plot(ride_time[threshold], ride_price[threshold], 
  cex=0.5,xlab = "Ride Time (Seconds)", ylab = "Ride Fare (Dollars)")
model <- lm(ride_price[threshold] ~ ride_time[threshold])
abline(model, col = "red")
estimators <- summary(model)$coefficients[, 4]
names(estimators) <- NULL
cat("The P-values for b0hat, b1hat are: ", estimators[1], 
    ", ", estimators[2], ", respectively.\n")
```

Since the p-values of the estimators are so low (at least lower than 0.025), we can reject the null hypothesis that the estimators are equal to 0 at a significance level of a = 0.05. This signifies a strong connection between length of the ride with price of the ride, which makes sense, since most taxis use a meter and charge by the minute.

## Plot 2 - Top 20 Taxi Companies

#### Bar Plot

This plot is a bar plot that shows us the top 20 taxi companies in Chicago.

```{r}
company <- (data[, .(Company)])
company_counts <- (company[, .N, by = Company])
ordered <- company_counts[order(-N)]
top20 <- head(ordered, 20)
names <- top20$Company
counts <- top20$N
par(mar = c(10, 4, 4, 2) + 0.1, tck = -0.01, xaxs = 'i', yaxs = 'i')
options(scipen = 999)
barplot(counts, 
        names.arg = names, 
        main = "Top 20 Taxi Companies", 
        ylab = "Number of Rides in 2024", 
        col = 'blue',
        las = 2,               
        cex.names = 0.7, 
        cex.axis = 0.7,
        ) 
abline(h = seq(0, max(counts), by = 200000), col = "grey80", lty = "dotted")
```

The plot shows that Flash Cab is the most popular cab service in Chicago with almost 1 million rides so far in 2024. Taxi Affiliation Services comes in second with about 750,000 rides this year and then Sun Taxi, and Taxicab Insurance Agency Llc are racing for the third position of number of taxi rides. We can also see the rest of the top 20 that have less and less uses.

## Plot 3 - Distance vs Tip Percentage

#### Line Plot

This plot is a line plot that shows the distance traveled by taxi vs the percentage that they get tipped.

```{r}
indexes <- sample(1:100000, 100000)
miles <- data$`Trip Miles`[indexes]
fare <- data$Fare[indexes]
tip <- data$Tips[indexes]
tipperc <- (tip / fare) * 100
df <- data.frame(miles, tipperc)
bin_width <- 0.5
bins <- cut(
  df$miles,
  breaks = seq(floor(min(df$miles, na.rm = TRUE)),ceiling(max(df$miles, na.rm = TRUE)),
    by = bin_width),
  include.lowest = TRUE,
  right = FALSE
)

tipbin <- tapply(df$tipperc, bins, mean, na.rm = TRUE)
bin_levels <- levels(bins)

centers <- sapply(strsplit(gsub("\\[|\\)|\\]", "", bin_levels), ","), function(x) {
  lower_bound <- as.numeric(x[1])
  lower_bound + (bin_width / 2)
})
valid_bins <- !is.na(tipbin)

centers <- centers[valid_bins]
tipbin <- tipbin[valid_bins]


sorted_indices <- order(centers)
centers_sorted <- centers[sorted_indices]
tipbin_sorted <- tipbin[sorted_indices]

plot(
  centers_sorted,
  tipbin_sorted,
  type = "l",  
  main = "Distance vs Tip Percentage",
  xlab = "Distance (Miles)",
  ylab = "Tip Percentage (%)",
  lwd = 2,
  col = "blue3",
  ylim = c(0, 25),
  xlim = c(0, 30)
)
grid()
points(centers_sorted, tipbin_sorted, pch = 19, col = "blue3")
```

This plot shows a very interesting dynamic between distance traveled and tip percentage. As you can see there is a sharp decline in percentage from 0 to 10 miles, and then a sharp increase from 11 to 15 miles and then a steady decline from there. We thought this was very interesting and hypothesized that this was due to the way that people tip. We believe that or shorter rides(0-10mi) people give a set tip not based on the fare, but on longer rides, tip is calculated by the fare.

## Plot 4 - Payment Types Used

#### Pie Chart

This plot shows the percentage distribution of payment types used for all taxi rides.

```{r}
ride_counts <- table(data$`Payment Type`)
ride_percentages <- prop.table(ride_counts) * 100
sorted_counts <- sort(ride_counts, decreasing = TRUE)
sorted_percentages <- prop.table(sorted_counts) * 100
distinct_colors <- c("lightblue", "orange", "purple", "green", "yellow", "red", "pink")
distinct_colors <- distinct_colors[1:length(sorted_counts)]
pie(sorted_counts, 
    labels = NA,  
    col = distinct_colors, 
    main = "Distribution of Rides by Payment Type (Percentage)")
legend("topright",  
       legend = paste(names(sorted_counts), ": ", round(sorted_percentages, 1), "%"),  
       fill = distinct_colors,  
       cex = 0.8) 
```

The pie chart shows that Credit Card payments dominate the taxi rides in Chicago, accounting for 39.7% of the total, followed by Cash payments at 28.6%. Mobile and PrCard payments collectively make up 27.3%, indicating significant usage of digital payment methods. Other categories, such as "Unknown" and "No Charge," represent much smaller percentages, highlighting the preference for more conventional payment methods like Credit Card and Cash.

## Plot 5 - Distribution of trip duration

#### Histogram

This plot shows a histogram of frequency of different trip durations(log-scaled).

```{r}
valid_durations <- data$`Trip Seconds`[data$`Trip Seconds` > 0 & 
  !is.na(data$`Trip Seconds`)]
log_durations <- log1p(valid_durations)
hist(log_durations, 
     breaks = 100,  
     col = "lightgreen", 
     main = "Log Distribution of Taxi Trip Durations in Chicago", 
     xlab = "Log(Trip Duration + 1)", 
     ylab = "Frequency of Rides", 
     border = "black")
lines(density(log_durations), col = "red", lwd = 2)
```

The graph shows the log-transformed distribution of taxi trip durations in Chicago, where the x-axis represents the logarithm of trip duration in seconds (with 1 added to handle zero values). Most taxi rides fall within the range of log values between 5 and 8, which corresponds to trip durations between approximately 150 seconds (2.5 minutes) and 3000 seconds (50 minutes). This indicates that the majority of taxi rides in Chicago are relatively short, typically ranging from a few minutes to under an hour. The density curve (in red) highlights the peak in the middle of the distribution, showing that trips around 15 to 30 minutes are the most common.

## Plot 6 - Trip duration at various times

#### Box Plot

This plot is a boxplot that shows how trip durations fluctuate at various times.

```{r}
data$trip_start_timestamp <- as.POSIXct(data$`Trip Start Timestamp`, 
  format="%m/%d/%Y %I:%M:%S %p")

data$hour <- as.numeric(format(data$trip_start_timestamp, "%I"))
data$ampm <- format(data$trip_start_timestamp, "%p")

data$hour_24 <- ifelse(data$ampm == "PM" & data$hour != 12, 
                       data$hour + 12, 
                       ifelse(data$ampm == "AM" & data$hour == 12, 0, data$hour))
data$time_of_day <- ifelse(data$hour_24 >= 6 & data$hour_24 < 12, "Morning",
                           ifelse(data$hour_24 >= 12 & data$hour_24 < 18, "Afternoon",
                                  ifelse(data$hour_24 >= 18 & data$hour_24 < 22, "Evening",
                                         ifelse(data$hour_24 >= 22 | data$hour_24 < 3,
        "Late Night", "Early Morning"
                                         ))))

filtered_data <- data[data$`Trip Seconds` <= 3000]

par(mar = c(7, 5, 4, 2))
boxplot(filtered_data$`Trip Seconds` ~ filtered_data$time_of_day,
        main = "Trip Duration (Seconds) by Time of Day (Trips Under 50 minutes)",
        xlab = "Time of Day",
        ylab = "Trip Duration (Seconds)",
        col = "lightblue", 
        las = 2,
        cex.axis = 0.6,
        cex.lab = 1.2,
        cex.main = 0.8)
```

It is interesting to note that the average trip time stayed mostly the same, with trips in the afternoon and morning being slightly shorter by a few minutes. In the early morning, IQR was the smallest, which we believed was due to using taxis to go to work in the morning, which is usually not far away, especially in the city. The biggest IQRs were in the afternoon and evening, where people use taxis for both short trips, like going home, and long trips, like going out to a restaurant or bar.
