
---
title: "Taxicab Project"
author: "Jack Dzialo, Tayten Bennetsen, Rith _____, Griffin Ball"
date: ''
format:
pdf: default
html: default
pdf-engine: pdflatex
editor: visual
fontsize: 10pt
geometry: margin=1in
toc: false 
toc-depth: 2 
---

# Introduction

Our team decided to analyze a dataset with information about taxi rides in Chicago to try and find links between taxi data such as ride time, price, and length with outside information such as weather conditions, different locations and their factors, and amount of traffic. The dataset contains over 5 million rows, and contains information of almost all taxi rides in the Chicago area from January 2024 to the present (October). It contains information such as but limited to: trip start and end times, trip length, fares and tips, company, and payment type.

# Plots

First, let us load in the data.

```{r}
library(data.table)
data <- data.table(fread("~/Downloads/taxi_chicago.csv"))
```

## Plot 1 - Scatterplot

This plot shows price of ride price vs. ride time.


```{r}
indexes <- sample(1:10000,1000)
ride_time <- data$`Trip Seconds`[indexes]
ride_price <- data$Fare[indexes]
outliers1 <- which(ride_price <= quantile(ride_price, .97, na.rm=TRUE))
outliers2 <- which(ride_time <= quantile(ride_time, .97, na.rm=TRUE))
threshold <- intersect(outliers1, outliers2)
plot(ride_time[threshold], ride_price[threshold], cex=0.5, xlab='Ride Time (Seconds)', ylab='Ride Fare (Dollars)')
model <- lm(ride_price[threshold] ~ ride_time[threshold])
abline(model, col='red')
estimators <- summary(model)$coefficients[,4]
names(estimators) <- NULL
cat('The P-values for b0hat, b1hat are: ',estimators[1], ', ', estimators[2], ', respectively.\n')
```

Since the p-values of the estimators are so low (at least lower than 0.025), we can reject the null hypothesis at a significance level of a = 0.05.

